Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-26 paper/main.tex
remove indent after equations in appendix

\section{Integrated Likelihood}
\label{sec:apdx_int_lik}
Here, we establish that under the model:

$$
\begin{aligned}
Y_i|S,W &\sim N\big(S, \Sigma_\varepsilon\big), \quad i \in \{1,\dots, r\}\\
S|W &\sim N\big(\boldsymbol{\mu}, \Sigma_S(w,w')\big)\\
W &\sim N\big(\mathbf{0}, \Sigma_W(x,x')\big) \\
\end{aligned}
$$

We can obtain the integrated likelihood $p(Y_i|W)=\int p(Y_i|S,W)p(S|W) dS$, 
with the following calculations:
$$
\begin{aligned}
p(Y_i|W) &= \int p(Y_i|S,W)p(S|W)dS \\
&\propto \int\exp\left(-\frac{1}{2} \left[Y_i^T\Sigma_\varepsilon^{-1} Y_i 
- 2Y_i^T\Sigma_\varepsilon^{-1} S + S^T \Sigma_\varepsilon^{-1} S + S^T \Sigma_S^{-1} S 
-2S^T \Sigma_S^{-1} \boldsymbol{\mu} 
+ \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right]\right)dS \\
\end{aligned}
$$

If we focus only on terms containing $S$, we have 
$$
\begin{aligned}
    \exp\Big(-\frac{1}{2} \big[S^T &(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
    -2S^T (\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \big]\Big) \\
&= \exp\left(-\frac{1}{2} \left[S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
-2S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \right]\right)
\end{aligned}
$$

To complete the square, we add and subtract $(\Sigma_S^{-1} \boldsymbol{\mu} 
+ \Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)$, 
which then gives the kernel for $S \sim N\left((\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i), 
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} \right)$. 
This removes the integral and leaves us with the following terms for $Y_i|W$:
$$
\begin{aligned}
&\exp\Big(-\frac{1}{2}\big[Y_i^T\Sigma_\varepsilon^{-1} Y + 
\boldsymbol{\mu}^T\Sigma_S^{-1} \boldsymbol{\mu} - (\Sigma_S^{-1} \boldsymbol{\mu} + 
\Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} 
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big[Y_i^T\big(\Sigma_\varepsilon^{-1} - 
\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1}\big)Y_i 
-2Y_i^T \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1} \boldsymbol{\mu} \\
&\qquad \qquad \qquad + \boldsymbol{\mu}^T (\Sigma_S^{-1} - \Sigma_S^{-1}(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) \boldsymbol{\mu}\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big((Y_i-\boldsymbol{\mu})^T(\Sigma_\varepsilon+\Sigma_S)^{-1}
(Y_i-\boldsymbol{\mu}) \big)\Big)
\end{aligned}
$$

With the last equality resulting from the use of the Sherman–Morrison–Woodbury formula, 
yielding $\big(\Sigma_\varepsilon^{-1} - \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1} = (\Sigma_S^{-1} - \Sigma_S^{-1}
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) = (\Sigma_\varepsilon+\Sigma_S)^{-1}$. 
We also use the fact that $\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_S^{-1} = (\Sigma_\varepsilon + \Sigma_S)^{-1}$, 
which holds since it can be shown that 
$[\Sigma_S(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})\Sigma_\varepsilon](\Sigma_\varepsilon+\Sigma_S)
=(\Sigma_\varepsilon+\Sigma_S)(\Sigma_\varepsilon+\Sigma_S)$.

@@

From our model specification, we have the following:

\[
S \sim \mathcal{N}(\boldsymbol{\mu}, \Sigma_S)
\]
\[
\bar{Y} \mid S \sim \mathcal{N}(S, \Sigma_\epsilon)
\]

Then, to find the distribution of $S|\bar{Y}$, we apply Bayes' Theorem below, 
where $\propto$ indicates proportionality:
\[
P(S \mid \bar{Y}) \propto P(\bar{Y} \mid S) P(S)
\]

\[
\propto \exp\left(-\tfrac{1}{2} (\bar{Y} - S)^T \Sigma_\epsilon^{-1} (\bar{Y} - S) \right) 
\cdot \exp\left(-\tfrac{1}{2} (S - \boldsymbol{\mu})^T \Sigma_S^{-1} (S - \boldsymbol{\mu})\right)
\]

\[
\propto \exp\left(-\tfrac{1}{2} \left[ \bar{Y}^T \Sigma_\epsilon^{-1} \bar{Y} 
- 2 S^T \Sigma_\epsilon^{-1} \bar{Y} + S^T \Sigma_\epsilon^{-1} S + S^T \Sigma_S^{-1} S 
- 2 S^T \Sigma_S^{-1} \boldsymbol{\mu} + \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right] \right)
\]

\[
\propto \exp\left( -\tfrac{1}{2} \left[ S^T (\Sigma_\epsilon^{-1} + \Sigma_S^{-1}) S 
- 2 S^T (\Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}) \right] \right)
\]

From properties of the multivariate normal distribution \citep[e.g., Equation 7.1 of][]{hoff2009first},
we can establish that the mean and covariance of $S|\bar{Y}$ will be $\boldsymbol{m}$
and $C$ respectively, where we can solve for these values with:

\[
C^{-1} = \Sigma_\epsilon^{-1} + \Sigma_S^{-1}
\]
\[
C^{-1} \boldsymbol{m} = \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}
\]

Then,
\[
C = (\Sigma_\epsilon^{-1} + \Sigma_S^{-1})^{-1}
\]
\[
\boldsymbol{m} = \left( \Sigma_S^{-1} + \Sigma_\epsilon^{-1} \right)^{-1} 
\left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
= C \cdot \left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
\]

Therefore, we have the result that $S|\bar{Y} \sim \mathcal{N}(\boldsymbol{m}, C)$.

>>>>

\section{Integrated Likelihood}
\label{sec:apdx_int_lik}
Here, we establish that under the model:
$$
\begin{aligned}
Y_i|S,W &\sim N\big(S, \Sigma_\varepsilon\big), \quad i \in \{1,\dots, r\}\\
S|W &\sim N\big(\boldsymbol{\mu}, \Sigma_S(w,w')\big)\\
W &\sim N\big(\mathbf{0}, \Sigma_W(x,x')\big) \\
\end{aligned}
$$
We can obtain the integrated likelihood $p(Y_i|W)=\int p(Y_i|S,W)p(S|W) dS$, 
with the following calculations:
$$
\begin{aligned}
p(Y_i|W) &= \int p(Y_i|S,W)p(S|W)dS \\
&\propto \int\exp\left(-\frac{1}{2} \left[Y_i^T\Sigma_\varepsilon^{-1} Y_i 
- 2Y_i^T\Sigma_\varepsilon^{-1} S + S^T \Sigma_\varepsilon^{-1} S + S^T \Sigma_S^{-1} S 
-2S^T \Sigma_S^{-1} \boldsymbol{\mu} 
+ \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right]\right)dS \\
\end{aligned}
$$
If we focus only on terms containing $S$, we have 
$$
\begin{aligned}
    \exp\Big(-\frac{1}{2} \big[S^T &(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
    -2S^T (\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \big]\Big) \\
&= \exp\left(-\frac{1}{2} \left[S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
-2S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \right]\right)
\end{aligned}
$$
To complete the square, we add and subtract $(\Sigma_S^{-1} \boldsymbol{\mu} 
+ \Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)$, 
which then gives the kernel for $S \sim N\left((\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i), 
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} \right)$. 
This removes the integral and leaves us with the following terms for $Y_i|W$:
$$
\begin{aligned}
&\exp\Big(-\frac{1}{2}\big[Y_i^T\Sigma_\varepsilon^{-1} Y + 
\boldsymbol{\mu}^T\Sigma_S^{-1} \boldsymbol{\mu} - (\Sigma_S^{-1} \boldsymbol{\mu} + 
\Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} 
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big[Y_i^T\big(\Sigma_\varepsilon^{-1} - 
\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1}\big)Y_i 
-2Y_i^T \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1} \boldsymbol{\mu} \\
&\qquad \qquad \qquad + \boldsymbol{\mu}^T (\Sigma_S^{-1} - \Sigma_S^{-1}(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) \boldsymbol{\mu}\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big((Y_i-\boldsymbol{\mu})^T(\Sigma_\varepsilon+\Sigma_S)^{-1}
(Y_i-\boldsymbol{\mu}) \big)\Big)
\end{aligned}
$$
With the last equality resulting from the use of the Sherman–Morrison–Woodbury formula, 
yielding $\big(\Sigma_\varepsilon^{-1} - \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1} = (\Sigma_S^{-1} - \Sigma_S^{-1}
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) = (\Sigma_\varepsilon+\Sigma_S)^{-1}$. 
We also use the fact that $\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_S^{-1} = (\Sigma_\varepsilon + \Sigma_S)^{-1}$, 
which holds since it can be shown that 
$[\Sigma_S(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})\Sigma_\varepsilon](\Sigma_\varepsilon+\Sigma_S)
=(\Sigma_\varepsilon+\Sigma_S)(\Sigma_\varepsilon+\Sigma_S)$.

@@

From our model specification, we have the following:
\[
S \sim \mathcal{N}(\boldsymbol{\mu}, \Sigma_S)
\]
\[
\bar{Y} \mid S \sim \mathcal{N}(S, \Sigma_\epsilon)
\]
Then, to find the distribution of $S|\bar{Y}$, we apply Bayes' Theorem below, 
where $\propto$ indicates proportionality:
\[
P(S \mid \bar{Y}) \propto P(\bar{Y} \mid S) P(S)
\]
\[
\propto \exp\left(-\tfrac{1}{2} (\bar{Y} - S)^T \Sigma_\epsilon^{-1} (\bar{Y} - S) \right) 
\cdot \exp\left(-\tfrac{1}{2} (S - \boldsymbol{\mu})^T \Sigma_S^{-1} (S - \boldsymbol{\mu})\right)
\]
\[
\propto \exp\left(-\tfrac{1}{2} \left[ \bar{Y}^T \Sigma_\epsilon^{-1} \bar{Y} 
- 2 S^T \Sigma_\epsilon^{-1} \bar{Y} + S^T \Sigma_\epsilon^{-1} S + S^T \Sigma_S^{-1} S 
- 2 S^T \Sigma_S^{-1} \boldsymbol{\mu} + \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right] \right)
\]
\[
\propto \exp\left( -\tfrac{1}{2} \left[ S^T (\Sigma_\epsilon^{-1} + \Sigma_S^{-1}) S 
- 2 S^T (\Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}) \right] \right)
\]
From properties of the multivariate normal distribution \citep[e.g., Equation 7.1 of][]{hoff2009first},
we can establish that the mean and covariance of $S|\bar{Y}$ will be $\boldsymbol{m}$
and $C$ respectively, where we can solve for these values with:
\[
C^{-1} = \Sigma_\epsilon^{-1} + \Sigma_S^{-1}
\]
\[
C^{-1} \boldsymbol{m} = \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}
\]
Then,
\[
C = (\Sigma_\epsilon^{-1} + \Sigma_S^{-1})^{-1}
\]
\[
\boldsymbol{m} = \left( \Sigma_S^{-1} + \Sigma_\epsilon^{-1} \right)^{-1} 
\left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
= C \cdot \left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
\]
Therefore, we have the result that $S|\bar{Y} \sim \mathcal{N}(\boldsymbol{m}, C)$.
